---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
terra::terraOptions(progress = 0)
```

# bsdesign: Biosecurity Surveillance and Area Freedom Design

<!-- badges: start -->
[![Last commit](https://img.shields.io/github/last-commit/cebra-analytics/bsdesign.svg)](https://github.com/cebra-analytics/bsdesign/commits/main)
<!-- badges: end -->

The *bsdesign* package provides a collection of workflow components implemented in *R* (<https://www.r-project.org/>) as *S3* object classes, which encapsulate functionality for building design models for effective allocations of biosecurity surveillance resources and area freedom analysis. Surveillance and area freedom design models are constructed by building and linking workflow components as follows:

1. Define the context of the biosecurity surveillance and/or area freedom design.
1. Specify the spatial or other divisions that the surveillance resources are to be allocated across.
1. Assemble a surveillance design model with configuration appropriate for the method and allocation conditions and generate the design.
1. Assemble an area freedom design model with configuration appropriate for the method and generate the design.

## Workflow components

This section further describes the workflow components for building surveillance and area freedom design models.

### Context

The context of the biosecurity surveillance and/or area freedom design is defined via the *Context* object class, and outlines information (where applicable and available) about the study including:

1. The threat species name.
1. The type of invasive species or threat:
    - Pest
    - Weed
    - Disease
1. The purpose of the surveillance:
    - Early detection of the threat
    - The delimitation of a threat incursion
    - Post-eradication monitoring of a threat
1. The type of surveillance utilised:
    - Surveys
    - Traps
    - Detectors
    - Samples
    - Reports (or general surveillance)
    - Mixed surveillance
1. The unit to describe surveillance resource quantities:
    - Units
    - Hours
    - Traps
    - Detectors
    - Samples
    - User specified units
1. The unit to describe costs/savings, including surveillance resource costs, incursion management costs, and/or surveillance benefit savings:
    - Dollars ($)
    - Hours
    - User specified cost units
1. The unit to describe surveillance time intervals when applicable:
    - Years
    - Months
    - Weeks
    - Days
    - Hours
    - User specified time units
1. The unit to describe spatial distances and areas (distance squared) when applicable:
    - Metres
    - Kilometres
    - User specified distance units
1. The status of the invasive species presence:
    - Never detected
    - Detected
    - Delimited
    - Contained
    - Eradicated
1. An indication of whether area freedom is declared.
1. An indication of whether absence or containment is required for market access.
1. The type of market access requirement:
    - Threat absence
    - Low prevalence of the threat
    - The threat is contained

### Divisions

The spatial or other divisions that the surveillance resources are to be allocated across are specified via the *Divisions* object class and may be configured as either:

1. A grid-based spatial raster layer (GeoTIFF file), whereby surveillance design parameters are specified, and allocations of surveillance resources are generated, across the active (non-NA) cell locations.
1. A network of spatial locations or patches defined via a table of longitude and latitude coordinates (CSV file), across which surveillance design parameters are specified, and allocations of surveillance resources are generated.
1. Non-spatial divisions (or spatial divisions without specified locations) defined via a table (CSV file), whereby surveillance design parameters are specified, and allocations of surveillance resources are generated, across divisions such as resource types, management categories, species, etc. 

### Surveillance designs

Surveillance designs may utilise various methods for generating effective surveillance resource allocations across specified *divisions*, and calculating surveillance sensitivities. The *SurveillanceDesign* object class provides a generic *base class* or template, with configuration and functionality common to different surveillance design approaches, including:

- Linking to a *Context* class object.
- Linking to a *Divisions* class object.
- Generating an effective surveillance resource allocation.
- Calculating the surveillance sensitivity (the probability of detection when the threat is present).
- Calculating the system-wide surveillance sensitivity across all divisions.
- Saving surveillance designs.

Several surveillance design approaches have been encapsulated in object classes that are based on (or *inherit*) the *SurveillanceDesign* class. Most generic functionality provided in the *SurveillanceDesign* class is overridden by inherited classes with implementations specific to the design method encapsulated. The following inherited object classes implement different design methods:

1. *AreaGrowthSurvDesign*: Implements a surveillance design method described in Epanchin-Niell et al. (2012) for the cost-effective allocation of surveillance sampling densities across one or more spatial sub-regions. The approach incorporates area-based population growth of one or more populations having discrete population size classes within each sub-region, and utilises
surveillance and other management costs, including costs associated with eradication, damages, and failure to contain incursions. The method implementation includes configuration for:
    - The surveillance context via a *Context* class object.
    - The divisions for specifying the sub-regions via a *Divisions* class object.
    - The area covered by each sub-region.
    - The expected number of populations to establish in each sub-region per time interval.
    - The area-based population growth model, including parameters for:
        - The area-based radial population growth rate in distance per time interval.
        - The maximum size class for populations represented in the growth model.
        - The maximum number of populations in a size class.
        - A function for calculating the area of a population given its size class ($s$) and the
growth rate ($g$), which by default is:\
        $f(g, s)=\pi\cdot(g\cdot s)^2$
    - The sample sensitivity for each sub-region.
    - Cost-based optimisation, including parameters for:
        - A list of estimated management costs for population incursions, including:
            - Eradication cost per area
            - Damage cost per area
            - Penalty cost per population of maximum size class
        - The cost per sample of allocated surveillance resources at each sub-region.
        - The cost budget or constraint (number of samples available) for the sampling allocation in the surveillance design.

2. *MultilevelSurvDesign*: Implements a multi-level sampling design methods described in Kean, Burnip, & Pathan (2015) for the effective allocation of surveillance sampling across one or more levels or stages, along with system-wide sensitivity calculation approaches outlined in Cannon (2009). The method implementation includes configuration for:
    - The surveillance context via a *Context* class object.
    - The divisions via a *Divisions* class object, specifying one or more multi-stage levels, ordered from lowest to highest (e.g. leaves, trees, rows, orchards). The lowest level can utilise either continuous density-based or discrete sampling. Higher levels can only utilise discrete sampling. Note that more than five levels may not be computationally feasible.
    - Calculating the overall sensitivity at each level (as per Kean, Burnip, & Pathan, 2015) via either:
        - Discrete sampling - whereby units are sampled at each level.
            - Sampling parameters include:
                - Sample sensitivity ($p$), or the probability of detection when the threat is present, for each level. The value is specified for the lowest level (only). Higher levels recursively utilise the calculated overall sensitivity from the previous lower level. 
                - Design prevalence ($P$) for each level.
                - The total number of sampling units ($N$) present at each level (e.g. leaves per tree, trees per row, rows per orchard).
            - The overall sensitivity $s(n)$ is calculated for a given allocation of samples ($n$) at each level via either:
                - $s(n) = 1 – (1 – p\cdot P)^{n}$\
                when up to 10% of the total units are sampled (i.e.$n \leq 0.1\cdot N$), or (assumed) when $N$ is unknown.
                - $s(n) = 1 – (1 – p\cdot n/N)^{P\cdot N}$\
                when more than 10% of the total units are sampled (i.e. $n > 0.1\cdot N$).
        - Continuous sampling - whereby a total area is sampled within the lowest level of the design.
            - Sampling parameters include:
                - Sample (area) sensitivity ($p$), or the probability of detection when the threat is present, for the lowest level (only).
                - Design density ($D$) for the lowest level (only).
                - Area of a single sample ($a$) at the lowest level (only). Note that when set to 1, the total number of samples ($n$) will be equivalent to the total area sampled ($A$).
            - The overall sensitivity $s(n)$ is calculated for a given total area sampled ($A$), or allocation of (area) samples ($n$), at the lowest level via either:
                - $s(n) = 1 – exp(–p\cdot A\cdot D)$\
                - $s(n) = 1 – exp(–p\cdot n\cdot a\cdot D)$
    - Finding an effective or optimal multi-level sampling design based on:
        - The cost of sampling at each level.
        - The desired (minimum) system-wide (highest level) sensitivity or detection probability of the multi-level sampling design (e.g. 0.95).

3. *RangeKernelSurvDesign*: Implements a surveillance design method described in Anderson et al. (2013; 2017; 2022) for the effective allocation of surveillance resources that are decreasingly effective across a spatial range, modelled via an exponential decay kernel. The functionality is extended with optimisation approaches, including those described in McCarthy et al. (2010) and Moore, McCarthy, & Lecomte (2016). The extended method implementation includes configuration for:
    - The surveillance context via a *Context* class object.
    - The divisions for specifying the spatial locations (grid-cells) for the design via a *Divisions* class object.
    - The establishment or occurrence probabilities of the threat at each spatial location.
    - Calculating the spatially distributed detection/capture probability of each allocated resource via:
        - The maximum detection or capture rates ($\lambda$) for each spatial location. Maximum detection/capture is achieved when a surveillance resource is placed at the centre of the home range of an invasive species individual or colony.
        - The spatial decay parameter ($\sigma$) for the (half-Normal) home range kernel. Note that $\sigma$ also specifies the maximum effective distance of detection/capture resources (i.e. twice the home range radius or $4\times \sigma$).
        - The number of time intervals ($t$) that each detection/capture resource or device is utilised (e.g. nights a trap is set).
        - An exponential (half-Normal) kernel for calculating detection/capture rates, which decays with distance ($d$):\
        $1 - (1 - \lambda\cdot exp(-d^2/(2\cdot \sigma^2)))^t$\
    - The cell-level design *prevalence* indicating the minimum number of location cells that are expected to be infected with the invasive species if the region of interest is infected. Values greater than 1 can be used as a proxy to population growth over time.
    - Finding an effective or optimal surveillance resource allocation design based on:
        - An optimisation strategy, either:
            - Maximum overall system-wide sensitivity
            - None for representing existing surveillance designs (see below)
        - Optimisation constraints, including:
            - The allocation budget or maximum number of surveillance resource units or devices
available.
            - The desired (minimum) system-wide sensitivity or detection probability of the surveillance design (e.g. 0.95).
    - Calculating sensitivities for existing sampling designs based on:
        - Existing surveillance resource quantities (e.g. samples or devices) at each location. Used when the optimisation strategy is specified as "none".
    - Detection sensitivities of existing surveillance present at each location may be optionally incorporated into the design.

4. *SamplingSurvDesign*: Implements an extended surveillance design method for the effective allocation of surveillance sampling across spatial or other divisions. The extended method is derived from sampling approaches described in Kean, Burnip, & Pathan (2015), sampling optimisation approaches described in Cannon (2009), along with generalised optimisation approaches described in Hauser & McCarthy (2009), McCarthy et al. (2010), and Moore, McCarthy, & Lecomte (2016), as well as approaches for incorporating existing detection sensitivities derived from those described in Anderson et al. (2017). The generalised optimisation functionality is encapsulated within the *LagrangeSurvDesign* object class (see below), which is utilised within both the *SamplingSurvDesign* and *SpatialSurvDesign* (see next section) object classes. The sampling surveillance design method includes configuration for:
    - The surveillance context via a *Context* class object.
    - The spatial locations or other divisions for the design via a *Divisions* class object.
    - The establishment or occurrence probabilities of the threat at each location or other division. Relative probabilities may be utilised.
    - Calculating the overall sensitivity at each location or other division as per Kean, Burnip, & Pathan (2015) via either:
        - Discrete sampling - whereby individual units are sampled at each location or other division.
            - Sampling parameters include:
                - Sample sensitivity ($p$), or the probability of detection when the threat is present, for each location or other division.
                - Design prevalence ($P$) for each location or other division. Note that this parameter may represent apparent prevalence (Cannon, 2009) when the sample sensitivity is set to 1.
                - The total number of individual sampling units ($N$), such as plants or animals, present at each location or other division.
            - The overall sensitivity $s(n)$ is calculated for a given allocation of samples ($n$) at a location or division via either:
                - $s(n) = 1 – (1 – p\cdot P)^{n}$\
                when up to 10% of the total units are sampled (i.e. $n \leq 0.1\cdot N$), or (assumed) when $N$ is unknown.
                - $s(n) = 1 – (1 – p\cdot n/N)^{P\cdot N}$\
                when more than 10% of the total units are sampled (i.e. $n > 0.1\cdot N$).
        - Continuous sampling - whereby a total area is sampled within each each location or other division.
            - Sampling parameters include:
                - Sample (area) sensitivity ($p$), or the probability of detection when the threat is present, for each location or other division.
                - Design density ($D$) for each location or other division.
                - Area of a single sample ($a$). Note that when set to 1, the total number of samples ($n$) will be equivalent to the total area sampled ($A$).
            - The overall sensitivity $s(n)$ is calculated for a given total area sampled ($A$), or allocation of (area) samples ($n$), within a location or division via either:
                - $s(n) = 1 – exp(–p\cdot A\cdot D)$\
                - $s(n) = 1 – exp(–p\cdot n\cdot a\cdot D)$
    - Finding an effective or optimal sampling design based on:
        - An optimisation strategy, either:
            - Minimum cost
            - Maximum saving
            - Maximum benefit
            - Maximum number of detections
            - Maximum overall system-wide sensitivity
            - None for representing existing surveillance designs (see below)
        - Optimisation parameters, including:
            - Estimated management costs for when the incursion is detected and undetected at each location or other division. Used for optimal minimum cost.
            - Estimated savings (cost-based benefit) at each location or other division. Used for optimal maximum saving.
            - Estimated (non-monetary) quantified benefit values at each location or other division. Used for optimal maximum benefit.
            - Cost per allocated sample at each location or other division.
            - Fixed costs, such as travel costs or time, at each location or other division.
        - Optimisation constraints, including:
            - The cost budget or maximum number of samples available.
            - The desired (minimum) system-wide sensitivity or detection probability of the surveillance design (e.g. 0.95).
            - The minimum permissible sampling allocation at each location or other division, to avoid impractically low sampling allocations.
            - An indication of whether the sampling allocation at each location or other division should be discrete integers (e.g. samples or devices), or continuous quantities (e.g. sampled areas or batches).
    - Calculating sensitivities for existing sampling designs based on:
        - Existing surveillance resource quantities (e.g. samples or devices) at each location or other division. Used when the optimisation strategy is specified as "none".
    - Detection sensitivities of existing surveillance present at each location or other division may be optionally incorporated into the design.

5. *SpatialSurvDesign*: Implements an extended surveillance design method for the effective allocation of surveillance resources across spatial (or other) divisions. The extended method is derived from optimisation approaches described in Hauser & McCarthy (2009), McCarthy et al. (2010), and Moore, McCarthy, & Lecomte (2016), along with approaches for incorporating existing detection sensitivities derived from those described in Anderson et al. (2017). The optimisation functionality is encapsulated within the *LagrangeSurvDesign* object class (see below), which is utilised within both the *SamplingSurvDesign* (see previous section) and *SpatialSurvDesign* object classes. The surveillance design method includes configuration for:
    - The surveillance context via a *Context* class object.
    - The spatial locations (or other divisions) for the design via a *Divisions* class object.
    - The establishment or occurrence probabilities of the threat at each location (or other division). Relative probabilities may be utilised.
    - Calculating the sensitivity, or probability of detecting an incursion when present, at each location (or other division) as per Hauser & McCarthy (2009), McCarthy et al. (2010), and Moore, McCarthy, & Lecomte (2016) via:
        - The efficacy ($\lambda$) or detection rates for each location (or other division).
        - The equation for calculating sensitivity $s(n)$ for a given allocation ($n$) of surveillance resources at a location (or other division):\
        $s(n) = 1 - exp(-\lambda\cdot n)$\
    - The cell-level design *prevalence* indicating the minimum number of location cells that are expected to be infected with the invasive species if the region of interest is infected. Values greater than 1 can be used as a proxy to population growth over time. Used when *divisions* specifies a cell-based (raster) region.
    - Finding an effective or optimal surveillance resource allocation design based on:
        - An optimisation strategy, either:
            - Minimum cost
            - Maximum saving
            - Maximum benefit
            - Maximum number of detections
            - Maximum overall system-wide sensitivity
            - None for representing existing surveillance designs (see below)
        - Optimisation parameters, including:
            - Estimated management costs for when the incursion is detected and undetected at each location (or other division). Used for optimal minimum cost.
            - Estimated savings (cost-based benefit) at each location (or other division). Used for optimal maximum saving.
            - Estimated (non-monetary) quantified benefit values at each location (or other division). Used for optimal maximum benefit.
            - Cost per unit of allocated surveillance resources at each location (or other division).
            - Fixed costs, such as travel costs or time, at each location (or other division).
        - Optimisation constraints, including:
            - The cost budget or maximum number of surveillance resources available.
            - The desired (minimum) system-wide sensitivity or detection probability of the surveillance design (e.g. 0.95).
            - The minimum permissible surveillance resource allocation at each location (or other division), to avoid impractically low resource allocations.
            - An indication of whether the resource allocation at each location (or other division) should be discrete integers (e.g. traps or detectors), or continuous quantities (e.g. survey hours).
    - Calculating sensitivities for existing surveillance designs based on:
        - Existing surveillance resource quantities (e.g. survey hours or devices) at each location (or other division). Used when the optimisation strategy is specified as "none".
    - Detection sensitivities of existing surveillance present at each location (or other division) may be optionally incorporated into the design.

#### Lagrange surveillance design optimisation

The *LagrangeSurvDesign* object class encapsulates an implementation of generalised optimisation approaches for the effective allocation of surveillance resources across one or more divisions (parts, locations, categories, etc.) via Lagrange-based methods described in Hauser & McCarthy (2009), McCarthy et al. (2010), and Moore, McCarthy, & Lecomte (2016). The implemented method is summarised by the following steps (see Moore, McCarthy, & Lecomte, 2016 - Appendix S3):

1. Formulate an objective function $f(n)$ for a given surveillance resource allocation ($n$) across locations or other divisions, based on:
    - The establishment or occurrence probabilities at each location/division.
    - The sensitivity formulation $s(n)$ (see previous sections).
    - An optimisation strategy, such as:
        - Minimum total costs
        - Maximum overall savings
        - Maximum benefit
        - Maximum number of detections
        - Maximum system-wide sensitivity
    - Detection/allocation costs, management costs, savings, or non-monetary benefits where applicable.
    - Criteria for avoiding negative allocations or allocations costing more than savings (where applicable), or other mathematical constraints (e.g. avoid dividing by zero), dependent on the formulation. The objective function is usually discontinuous due to criteria.\
    - The objective function is commonly a summation across locations/divisions ($i$):\
    $f(n) = \sum_if_i(n_i)$\
    Except for the objective function for maximum system-wide sensitivity ($s_{system}$) where:\
    $s_{system}(n) = 1 - \prod_i(1-s(n_i))$ is utilised in the formulation of $f(n)$.
2. Derive the (partial) derivative $f'(n)$ (or marginal benefit) of the objective function. The optimal solution occurs when the marginal benefit is constant ($\alpha$), that is:\
    $f_i'(n_i) = \alpha$ for all locations/divisions ($i$).
3. Derive the pseudo-inverse $f^+(\alpha)$ of the derivative function via algebraic rearrangement of the derivative (if possible), such that given an *alpha* value ($\alpha$) the function generates allocations ($n_i$) for each location/division ($i$).
4. Perform an iterative search for the *alpha* value that corresponds to the optimal solution for the objective function $f(n)$, given any budget or other constraints (e.g. desired minimum system-wide sensitivity). Note that budget or other constraints are optional for cost or saving-based optimisation.

The *LagrangeSurvDesign* object class is utilise within both the *SamplingSurvDesign* and *SpatialSurvDesign* object classes (see previous sections), and includes configuration for:

- The surveillance context via a *Context* class object.
- The spatial locations or other divisions for the design via a *Divisions* class object.
- The establishment or occurrence probabilities of the threat at each location or other division. Relative probabilities may be utilised.
- The formulated functions (as above) specific to configured surveillance designs (within *SamplingSurvDesign* and *SpatialSurvDesign* object classes), including:
    - The objective function $f(n)$.
    - The (partial) derivative $f'(n)$ of the objective function.
    - The pseudo-inverse $f^+(\alpha)$ of the derivative function, plus configuration for $\alpha$ value constraints.
- Functions for calculating sensitivity $s_i(n_i)$ and its inverse (i.e. allocations) $n_i(s_i)$ at each location/division ($i$).
- Optimisation constraints:
    - The cost budget or maximum number of surveillance resources available.
    - The desired (minimum) system-wide sensitivity or detection probability of the surveillance design (e.g. 0.95).
    - Minimum allocation quantities at each location/division.

### Area freedom designs 

(coming soon)

## Installation

You can install the latest version of *bsdesign* from [GitHub](https://github.com/) with:

``` r
remotes::install_github("cebra-analytics/bsdesign")
```

## Surveillance example

The following example builds a surveillance design for Orange Hawkweed (*Hieracium aurantiacum*), an exotic weed for Australia, which approximately reproduces the surveillance design described in Hauser & McCarthy (2009). In the following steps we build the workflow components, run the design model, and examine the surveillance design results.

### Step 1: Context

(coming soon)

### Step 2: Divisions

In our example we will allocate surveillance resources across grid-based spatial divisions. The region of interest for our Hawkweed spread model is the Bogong High Plains area surrounding the ski-resort township of Falls Creek in Victoria, Australia. We derive template raster (GeoTIFF) for the region from the National Vegetation Information System (NVIS) V7.0 (NVIS, 2025) raster layers. We also use the NVIS data in step 2. Users may download [NVIS Raster Geodatabases](https://www.dcceew.gov.au/environment/environment-information-australia/national-vegetation-information-system/data-products) and place them in a suitable directory (e.g. *downloaded_data*) before loading and transforming the NVIS layer. We build our *Divisions* class object with our 100 m resolution template.

```{r example_2, message = FALSE, fig.align = "center"}
# Load NVIS major vegetation groups (MGV)
nvis_rast <- terra::rast(
  paste0("../downloaded_data/NVIS_V7_0_AUST_RASTERS_EXT_ALL/",
         "NVIS_V7_0_AUST_EXT.gdb"))[["NVIS7_0_AUST_EXT_MVG_ALB"]]
# Crop to the region of interest
region_nvis_rast <- terra::crop(nvis_rast,
                                c(1354000, 1372000, -4120000, -4103000))
# Divisions class object configured with a masked NVIS template
divisions <- bsdesign::Divisions(+(region_nvis_rast > 0))
terra::plot(divisions$get_rast(1), colNA = "grey",
            main = "Hawkweed example grid-based divisions")
```

### Step 3: Surveillance design

(coming soon)

## Area freedom example

(coming soon)

## References

Anderson, D. P., Gormley, A. M., Ramsey, D. S. L., Nugent, G., Martin, P. A. J., Bosson, M., Livingstone, P., & Byrom, A. E. (2017). 'Bio-economic optimisation of surveillance to confirm
broadscale eradications of invasive pests and diseases'. *Biological Invasions*, 19(10), 2869–2884. [doi:10.1007/s1053001714905](https://doi.org/10.1007/s1053001714905)

Anderson, D. P., Pepper, M. A., Travers, S., Michaels, T. A., Sullivan, K., & Ramsey, D. S. L.
(2022). 'Confirming the broadscale eradication success of nutria (Myocastor coypus) from the Delmarva Peninsula, USA'. *Biological Invasions*. [doi:10.1007/s1053002202855x](https://doi.org/10.1007/s1053002202855x)

Anderson, D. P., Ramsey, D. S. L., Nugent, G., Bosson, M., Livingstone, P., Martin, P. A. J.,
Sergeant, E., Gormley, A. M., & Warburton, B. (2013). 'A novel approach to assess the probability
of disease eradication from a wild-animal reservoir host'. *Epidemiology and Infection*, 141(7),
1509–1521. [doi:10.1017/S095026881200310X](https://doi.org/10.1017/S095026881200310X)

Barclay, H. J., & Hargrove, J. W. (2005). 'Probability models to facilitate a declaration of pest-free status, with special reference to tsetse (Diptera: Glossinidae)'. *Bulletin Of Entomological Research (London)*, 95(1), 1–12. [doi:10.1079/BER2004331](https://doi.org/10.1079/BER2004331)

Cannon, R. M. (2009). 'Inspecting and monitoring on a restricted budget - where best to look?'
*Preventive Veterinary Medicine*, 92(1–2), 163-174. [doi:10.1016/j.prevetmed.2009.06.009](https://doi.org/10.1016/j.prevetmed.2009.06.009)

CRC (2003). Orange Hawkweed Weed Management Guide (2003) CRC for Weed Management. Accessed in 2026 via <https://weeds.org.au/profiles/orange-hawkweed/>

Epanchin-Niell, R. S., Haight, R. G., Berec, L., Kean, J. M., & Liebhold, A. M. (2012). 'Optimal
surveillance and eradication of invasive species in heterogeneous landscapes'. *Ecology Letters*,
15(8), 803–812. [doi:10.1111/j.14610248.2012.01800.x](https://doi.org/10.1111/j.14610248.2012.01800.x]

Hauser, C. E., Giljohann, K. M., Rigby, M., Herbert, K., Curran, I., Pascoe, C., Williams, N. S. G., Cousens, R. D., & Moore, J. L. (2016). 'Practicable methods for delimiting a plant invasion'. *Diversity and Distributions*, 22(1/2), 136–147. [doi:10.1111/ddi.12388](https://doi.org/10.1111/ddi.12388)

Hauser, C. E., & McCarthy, M. A. (2009). 'Streamlining "search and destroy": cost-effective surveillance for invasive species management'. *Ecology Letters*, 12(7), 683–692. [doi:10.1111/j.1461-0248.2009.01323.x](https://doi.org/10.1111/j.1461-0248.2009.01323.x)

Kean, J. M., Burnip, G. M. & Pathan, A. (2015). Detection survey design for decision making during biosecurity incursions. In F. Jarrad, S. Low-Choy & K. Mengersen (eds.), *Biosecurity surveillance: Quantitative approaches* (pp. 238– 250). Wallingford, UK: CABI. [doi:10.1079/9781780643595.0000](https://doi.org/10.1079/9781780643595.0000)

Magarey, R. C., Reynolds, M., Dominiak, B. C., Sergeant, E., Agnew, J.,Ward, A., & Thompson, N.
(2019). 'Review of sugarcane Fiji leaf gall disease in Australia and the declaration of pest freedom in Central Queensland'. *Crop Protection*, 121, 113–120. [doi:10.1016/j.cropro.2019.03.022](https://doi.org/10.1016/j.cropro.2019.03.022)

McCarthy, M. A., Thompson, C. J., Hauser, C., Burgman, M. A., Possingham, H. P., Moir, M. L.,
Tiensin, T., & Gilbert, M. (2010). 'Resource allocation for efficient environmental management'.
*Ecology Letters*, 13(10), 1280–1289. [doi:10.1111/j.14610248.2010.01522.x](https://doi.org/10.1111/j.14610248.2010.01522.x)

Moore, A. L., McCarthy, M. A., & Lecomte, N. (2016). 'Optimizing ecological survey effort over
space and time'. *Methods in Ecology and Evolution*, 7(8), 891–899. [doi:10.1111/2041210X.12564](https://doi.org/10.1111/2041210X.12564)

National Vegetation Information System (NVIS) V7.0 <https://www.dcceew.gov.au/> from copyright Commonwealth of Australia 2025

Regan, T. J., McCarthy, M. A., Baxter, P. W., Panetta, F. D., & Possingham, H. P. (2006). 'Optimal eradication: when to stop looking for an invasive plant'. *Ecology Letters*, 9(7), 759–766.
[doi:10.1111/j.14610248.2006.00920.x](https://doi.org/10.1111/j.14610248.2006.00920.x)

Rout, T. (2017). 'Declaring Eradication of an Invasive Species'. In A. Robinson, T. Walshe, M.
Burgman, & M. Nunn (Eds.), *Invasive Species: Risk Assessment and Management* (pp. 334-347).
Cambridge: Cambridge University Press. [doi:10.1017/9781139019606.017](https://doi.org/10.1017/9781139019606.017)

Solow, A. R. (1993). 'Inferring Extinction from Sighting Data'. *Ecology*, 74(3), 962–964. [doi:10.2307/1940821](https://doi.org/10.2307/1940821)

Williams, N. S. G., Hahs, A. K., & Morgan, J. W. (2008). 'A Dispersal-Constrained Habitat Suitability Model for Predicting Invasion of Alpine Vegetation'. *Ecological Applications*, 18(2), 347–359. [doi:10.1890/07-0868.1](https://doi.org/10.1890/07-0868.1)
