---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
terra::terraOptions(progress = 0)
```

# bsdesign: Biosecurity Surveillance and Area Freedom Design

<!-- badges: start -->
[![Last commit](https://img.shields.io/github/last-commit/cebra-analytics/bsdesign.svg)](https://github.com/cebra-analytics/bsdesign/commits/main)
<!-- badges: end -->

The *bsdesign* package provides a collection of workflow components implemented in *R* (<https://www.r-project.org/>) as *S3* object classes, which encapsulate functionality for building design models for effective allocations of biosecurity surveillance resources and area freedom analysis. Surveillance and area freedom design models are constructed by building and linking workflow components as follows:

1. Define the context of the biosecurity surveillance and/or area freedom design.
1. Specify the spatial or other divisions that the surveillance resources are to be allocated across.
1. Assemble a surveillance design model with configuration appropriate for the method and allocation conditions and generate the design.
1. Assemble an area freedom design model with configuration appropriate for the method and generate the design.

## Workflow components

This section further describes the workflow components for building surveillance and area freedom design models.

### Context

The context of the biosecurity surveillance and/or area freedom design is defined via the *Context* object class, and outlines information about the study including:

1. The threat species name.
1. The type of invasive species or threat:
    - Pest
    - Weed
    - Disease
1. The purpose of the surveillance:
    - Early detection of the threat
    - The delimitation of a threat incursion
    - Post-eradication monitoring of a threat
1. The type of surveillance utilised:
    - Surveys
    - Traps
    - Detectors
    - Samples
    - Reports (or general surveillance)
    - Mixed surveillance
1. The unit to describe surveillance resource quantities:
    - Units
    - Hours
    - Traps
    - Detectors
    - Samples
    - User specified units
1. The unit to describe costs/savings, including surveillance resource costs, incursion management costs, and/or surveillance benefit savings:
    - Dollars ($)
    - Hours
    - User specified cost units
1. The unit to describe surveillance time intervals when applicable:
    - Years
    - Months
    - Weeks
    - Days
    - Hours
    - User specified time units
1. The unit to describe spatial distances and areas (distance squared) when applicable:
    - Metres
    - Kilometres
    - User specified distance units
1. The status of the invasive species presence:
    - Never detected
    - Detected
    - Delimited
    - Contained
    - Eradicated
1. An indication of whether area freedom is declared.
1. An indication of whether absence or containment is required for market access.
1. The type of market access requirement:
    - Threat absence
    - Low prevalence of the threat
    - The threat is contained

### Divisions

The spatial or other divisions that the surveillance resources are to be allocated across are specified via the *Divisions* object class and may be configured as either:

1. A grid-based spatial raster layer (GeoTIFF file), whereby surveillance design parameters are specified, and allocations of surveillance resources are generated, across the active (non-NA) cell locations.
1. A network of spatial locations or patches defined via a table of longitude and latitude coordinates (CSV file), across which surveillance design parameters are specified, and allocations of surveillance resources are generated.
1. Non-spatial divisions (or spatial divisions without specified locations) defined via a table (CSV file), whereby surveillance design parameters are specified, and allocations of surveillance resources are generated, across divisions such as resource types, management categories, species, etc. 

### Surveillance designs

Surveillance designs may utilise various methods for generating effective surveillance resource allocations across specified *divisions*, and calculating surveillance sensitivities. The *SurveillanceDesign* object class provides a generic *base class* or template, with configuration and functionality common to different surveillance design approaches, including:

- Linking to a *Context* class object.
- Linking to a *Divisions* class object.
- Generating an effective surveillance resource allocation.
- Calculating the surveillance sensitivity (the probability of detection when the threat is present).
- Calculating the system-wide surveillance sensitivity across all divisions.
- Saving surveillance designs.

Several surveillance design approaches have been encapsulated in object classes that are based on (or *inherit*) the *SurveillanceDesign* class. Most generic functionality provided in the *SurveillanceDesign* class is overridden by *inherited* classes with implementations specific to the design method encapsulated. The following *inherited* object classes implement different design methods:

1. *AreaGrowthSurvDesign*: Implements a surveillance design method described in Epanchin-Niell et al. (2012) for the cost-effective allocation of surveillance sampling densities across one or more spatial sub-regions. The approach incorporates area-based population growth of one or more populations (having discrete population size classes) within each sub-region, and utilises
surveillance and other management costs, including costs associated with eradication, damages, and failure to contain incursions. The method implementation includes configuration for:
    - The surveillance *context* via a *Context* class object.
    - The *divisions* for specifying the sub-regions via a *Divisions* class object.
    - The area covered by each sub-region.
    - The expected number of populations to establish in each sub-region per time interval (*establish rate*).
    - The area-based radial population *growth rate* in distance per time interval.
    - The maximum size class for populations represented in the growth model.
    - The maximum number of populations in a size class.
    - A function for calculating the area of a population given its size class and the
growth rate, which by default is:\
    $f(growth\_rate, size\_class)=\pi\cdot(growth\_rate\cdot size\_class)^2$
    - The sample sensitivity for each sub-region.
    - A list of estimated management costs (*mgmt cost*) for population incursions, including:
        - Eradication cost per area
        - Damage cost per area
        - Penalty cost per population of maximum size class
    - The cost per sample of allocated surveillance resources at each sub-region.
    - The cost *budget* or constraint (number of samples available) for the sampling allocation in the surveillance design.

2. *MultilevelSurvDesign*: Implements a multi-level sampling design methods described in Kean, Burnip, & Pathan (2015) for the effective allocation of surveillance sampling across one or more levels or stages, along with system-wide sensitivity calculation approaches outlined in Cannon (2009). The method implementation includes configuration for:
    - The surveillance *context* via a *Context* class object.
    - The *divisions* (a *Divisions* class object) for specifying one or more multi-stage levels, ordered from lowest to highest (e.g. leaves, trees, rows, orchards). The lowest level can utilise either *continuous* density-based or *discrete* sampling. Higher levels can only utilise *discrete* sampling. Note that more than five levels may not be computationally feasible.
    - A single sample sensitivity value for the lowest level of the multi-level/stage sampling design.
    - The type of sampling used for the lowest level of the multilevel sampling, either:
        - Discrete
        - Continuous
    - The sampling design *prevalence* for each multi-level/stage. Note that this parameter may represent apparent prevalence (Cannon, 2009) when the sensitivity is set to 1.
    - The total number of individual *discrete* sampling units (e.g. leaves, trees, rows,orchards) present at each multi-stage level.
    - The *design density* for when the lowest level of the multilevel sampling is *continuous*.
    - The area of a single sample when the lowest level for when the lowest level of the multi-level sampling is *continuous*. Note that when set to 1, the total number of samples will be equivalent to the total area sampled.
    - The cost of samples at each level. An attribute *units* may be used to specify the cost units (e.g. "$" or "hours").
    - The desired (minimum) system-wide sensitivity or detection probability of the surveillance
design (e.g. 0.95).

3. *RangeKernelSurvDesign*: Implements a surveillance design method described in Anderson et al. (2013; 2017; 2022) for the effective allocation of surveillance resources that are decreasingly effective across a spatial range, modelled via an exponential decay kernel. The functionality is extended with optimisation approaches, including those described in McCarthy et al. (2010) and Moore, McCarthy, & Lecomte (2016). The extended method implementation includes configuration for:
    - The surveillance *context* via a *Context* class object.
    - The *divisions* for specifying the spatial locations (grid-cells) for the design via a *Divisions* class object.
    - Establishment or occurrence probabilities of the threat at each spatial location. Values are assumed to be relative when their maximum is greater than 1, or an attribute *relative* is attached to the parameter and set to *TRUE*.
    - Maximum detection or capture rates (*lambda*) for each spatial location. Maximum detection/capture is achieved when a surveillance resource is placed in the centre of the home range of an invasive species individual or colony. Detection/capture rate decays with distance via an exponential kernel such that the detection/capture probability of each allocated resource can be expressed via:\
    $1 - (1 - lambda\cdot exp(\frac{-distance^2}{2\cdot sigma^2})^{intervals})$\
    for a given number of time *intervals* and up to a maximum distance of 4 $\times$ *sigma*, where *sigma* is the home range decay parameter.
    - The spatial decay parameter (*sigma*) for the (half-Normal) home range kernel (above). Note that *sigma* also specifies the maximum effective distance of detection/capture resources (i.e. twice the home range radius or 4 $\times$ *sigma*).

4. *SamplingSurvDesign*:

(in progress)

5. *SpatialSurvDesign*:

(in progress)

### Area freedom designs 

(coming soon)

## Installation

You can install the latest version of *bsdesign* from [GitHub](https://github.com/) with:

``` r
remotes::install_github("cebra-analytics/bsdesign")
```

## Surveillance example

The following example builds a surveillance design for Orange Hawkweed (*Hieracium aurantiacum*), an exotic weed for Australia, which approximately reproduces the surveillance design described in Hauser & McCarthy (2009). In the following steps we build the workflow components, run the design model, and examine the surveillance design results.

### Step 1: Context

(coming soon)

### Step 2: Divisions

In our example we will allocate surveillance resources across grid-based spatial divisions. The region of interest for our Hawkweed spread model is the Bogong High Plains area surrounding the ski-resort township of Falls Creek in Victoria, Australia. We derive template raster (GeoTIFF) for the region from the National Vegetation Information System (NVIS) V7.0 (NVIS, 2025) raster layers. We also use the NVIS data in step 2. Users may download [NVIS Raster Geodatabases](https://www.dcceew.gov.au/environment/environment-information-australia/national-vegetation-information-system/data-products) and place them in a suitable directory (e.g. *downloaded_data*) before loading and transforming the NVIS layer. We build our *Divisions* class object with our 100 m resolution template.

```{r example_2, message = FALSE, fig.align = "center"}
# Load NVIS major vegetation groups (MGV)
nvis_rast <- terra::rast(
  paste0("../downloaded_data/NVIS_V7_0_AUST_RASTERS_EXT_ALL/",
         "NVIS_V7_0_AUST_EXT.gdb"))[["NVIS7_0_AUST_EXT_MVG_ALB"]]
# Crop to the region of interest
region_nvis_rast <- terra::crop(nvis_rast,
                                c(1354000, 1372000, -4120000, -4103000))
# Divisions class object configured with a masked NVIS template
divisions <- bsdesign::Divisions(+(region_nvis_rast > 0))
terra::plot(divisions$get_rast(1), colNA = "grey",
            main = "Hawkweed example grid-based divisions")
```

### Step 3: Surveillance design

(coming soon)

### Area freedom example

(coming soon)

## References

Anderson, D. P., Gormley, A. M., Ramsey, D. S. L., Nugent, G., Martin, P. A. J., Bosson, M., Livingstone, P., & Byrom, A. E. (2017). 'Bio-economic optimisation of surveillance to confirm
broadscale eradications of invasive pests and diseases'. *Biological Invasions*, 19(10), 2869–2884. [doi:10.1007/s1053001714905](https://doi.org/10.1007/s1053001714905)

Anderson, D. P., Pepper, M. A., Travers, S., Michaels, T. A., Sullivan, K., & Ramsey, D. S. L.
(2022). 'Confirming the broadscale eradication success of nutria (Myocastor coypus) from the Delmarva Peninsula, USA'. *Biological Invasions*. [doi:10.1007/s1053002202855x](https://doi.org/10.1007/s1053002202855x)

Anderson, D. P., Ramsey, D. S. L., Nugent, G., Bosson, M., Livingstone, P., Martin, P. A. J.,
Sergeant, E., Gormley, A. M., & Warburton, B. (2013). 'A novel approach to assess the probability
of disease eradication from a wild-animal reservoir host'. *Epidemiology and Infection*, 141(7),
1509–1521. [doi:10.1017/S095026881200310X](https://doi.org/10.1017/S095026881200310X)

Barclay, H. J., & Hargrove, J. W. (2005). 'Probability models to facilitate a declaration of pest-free status, with special reference to tsetse (Diptera: Glossinidae)'. *Bulletin Of Entomological Research (London)*, 95(1), 1–12. [doi:10.1079/BER2004331](https://doi.org/10.1079/BER2004331)

Cannon, R. M. (2009). 'Inspecting and monitoring on a restricted budget - where best to look?'
*Preventive Veterinary Medicine*, 92(1–2), 163-174. [doi:10.1016/j.prevetmed.2009.06.009](https://doi.org/10.1016/j.prevetmed.2009.06.009)

CRC (2003). Orange Hawkweed Weed Management Guide (2003) CRC for Weed Management. Accessed in 2026 via <https://weeds.org.au/profiles/orange-hawkweed/>

Epanchin-Niell, R. S., Haight, R. G., Berec, L., Kean, J. M., & Liebhold, A. M. (2012). 'Optimal
surveillance and eradication of invasive species in heterogeneous landscapes'. *Ecology Letters*,
15(8), 803–812. [doi:10.1111/j.14610248.2012.01800.x](https://doi.org/10.1111/j.14610248.2012.01800.x]

Hauser, C. E., Giljohann, K. M., Rigby, M., Herbert, K., Curran, I., Pascoe, C., Williams, N. S. G., Cousens, R. D., & Moore, J. L. (2016). 'Practicable methods for delimiting a plant invasion'. *Diversity and Distributions*, 22(1/2), 136–147. [doi:10.1111/ddi.12388](https://doi.org/10.1111/ddi.12388)

Hauser, C. E., & McCarthy, M. A. (2009). 'Streamlining "search and destroy": cost-effective surveillance for invasive species management'. *Ecology Letters*, 12(7), 683–692. [doi:10.1111/j.1461-0248.2009.01323.x](https://doi.org/10.1111/j.1461-0248.2009.01323.x)

Kean, J. M., Burnip, G. M. & Pathan, A. (2015). Detection survey design for decision making during biosecurity incursions. In F. Jarrad, S. Low-Choy & K. Mengersen (eds.), *Biosecurity surveillance: Quantitative approaches* (pp. 238– 250). Wallingford, UK: CABI. [doi:10.1079/9781780643595.0000](https://doi.org/10.1079/9781780643595.0000)

Magarey, R. C., Reynolds, M., Dominiak, B. C., Sergeant, E., Agnew, J.,Ward, A., & Thompson, N.
(2019). 'Review of sugarcane Fiji leaf gall disease in Australia and the declaration of pest freedom in Central Queensland'. *Crop Protection*, 121, 113–120. [doi:10.1016/j.cropro.2019.03.022](https://doi.org/10.1016/j.cropro.2019.03.022)

McCarthy, M. A., Thompson, C. J., Hauser, C., Burgman, M. A., Possingham, H. P., Moir, M. L.,
Tiensin, T., & Gilbert, M. (2010). 'Resource allocation for efficient environmental management'.
*Ecology Letters*, 13(10), 1280–1289. [doi:10.1111/j.14610248.2010.01522.x](https://doi.org/10.1111/j.14610248.2010.01522.x)

Moore, A. L., McCarthy, M. A., & Lecomte, N. (2016). 'Optimizing ecological survey effort over
space and time'. *Methods in Ecology and Evolution*, 7(8), 891–899. [doi:10.1111/2041210X.12564](https://doi.org/10.1111/2041210X.12564)

National Vegetation Information System (NVIS) V7.0 <https://www.dcceew.gov.au/> from copyright Commonwealth of Australia 2025

Regan, T. J., McCarthy, M. A., Baxter, P. W., Panetta, F. D., & Possingham, H. P. (2006). 'Optimal eradication: when to stop looking for an invasive plant'. *Ecology Letters*, 9(7), 759–766.
[doi:10.1111/j.14610248.2006.00920.x](https://doi.org/10.1111/j.14610248.2006.00920.x)

Rout, T. (2017). 'Declaring Eradication of an Invasive Species'. In A. Robinson, T. Walshe, M.
Burgman, & M. Nunn (Eds.), *Invasive Species: Risk Assessment and Management* (pp. 334-347).
Cambridge: Cambridge University Press. [doi:10.1017/9781139019606.017](https://doi.org/10.1017/9781139019606.017)

Solow, A. R. (1993). 'Inferring Extinction from Sighting Data'. *Ecology*, 74(3), 962–964. [doi:10.2307/1940821](https://doi.org/10.2307/1940821)

Williams, N. S. G., Hahs, A. K., & Morgan, J. W. (2008). 'A Dispersal-Constrained Habitat Suitability Model for Predicting Invasion of Alpine Vegetation'. *Ecological Applications*, 18(2), 347–359. [doi:10.1890/07-0868.1](https://doi.org/10.1890/07-0868.1)
